---
title: "Supplemental Figure"
author: "Abby"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  pdf_document:
    latex_engine: xelatex
  dev: pdf

---

#Reviewer comment #1

Figure S2 monte carlo test sounds strange for what the authors want to test (page 4 top). If the idea is to test whether the day before was the most similar, why not just test it directly and compare with distance to other days, rather than this permutation test? Am I missing something? And why is Bray-Curtis distance used here?

#Reviewer comment #2

The authors state that "each subject's microbiome samples were most similar to their own sample from the previous day compared to all other timepoints from that subject". Was this done also with comparing to all other unrelated samples, or only to samples within the same subject? The proper analysis should be to compare as background to the whole cohort.


# To address

The analysis should be redone using the Aitchison's distance.Get each day's distance to the next day within in a person and compare to all of the distances that are 2 or more days apart.

Then test whether the mean difference is <0 with a t-test. 

```{r setup, include=FALSE, message = FALSE}

require(rmarkdown)
require(knitr)
require(ggplot2)
require(reshape2)

# set the path for root directory
opts_knit$set(root.dir = "/Users/abby/Documents/Projects/dietstudy/")

# set path for figures
opts_chunk$set(echo = TRUE, fig.path = "../output/Figs/", dev = c("pdf"))

```

```{r load and organize data, include = F, message = F}
# load map
map <- read.delim(file = "data/maps/SampleID_map.txt")
map <- map[colnames(map) %in% c("X.SampleID", "UserName","StudyDayNo","Sample.Week.Day")]

# load mb table for everyone
mb <- read.delim("data/processed_tax/taxonomy_clr_s.txt", row = 1)

# get mb distances (euclidean from clr)
mbdist <- dist(t(mb))
mbdist <- as.matrix(mbdist)

#limit map to the people with mb samples
map <- map[map$X.SampleID %in% colnames(mb),]
map <- droplevels(map)

# listify distance matrices by person
list.mbdist<- NULL

for (i in seq_along(unique(map$UserName))){
  a <- as.character(unique(map$UserName)[i])
  ids <- droplevels(map$X.SampleID[map$UserName == a])
  
  sub.mbdist <- mbdist[rownames(mbdist) %in% ids,colnames(mbdist) %in% ids]
  list.mbdist[[i]] <- sub.mbdist
  
} 


# calculate distance to the next day and distance to days 2 days apart

### This is the meat of the problem, and where I think I'm probably messing up! ###

day1diff <- NULL
day1diffall <- NULL
weekdaysall <- NULL
allotherdaysdiff <- NULL
for (i in 1:length(list.mbdist)) {
  # subset to just one person 
  x <- list.mbdist[[i]]
  n <- dim(x)[1]
  # pull the distances between sucessive days
  # create an indicator for the diagonals in the matrix
  d <- row(x) - col(x)
  # use split to group on these values by their diagonals 
  vals <- split(x, d)
  # get the mean of the diagonal that corresponds to 1 day diff
  day1diff[[i]] <- mean(unlist(vals[n+1]))
  day1diffall[[i]] <- unlist(vals[n+1])
  # get the mean of all the other distances
  allotherdaysdiff[[i]] <- mean(unlist(vals[(n+2):(2*n-1)]))
  
  # get the days for plotting
  days <- colnames(x)[2:length(colnames(x))]
  w <- subset(map, map$X.SampleID %in% days)
  weekdaysall[[i]] <- as.vector(w[,"StudyDayNo"])
  
}

x <- unlist(day1diff)
y <- unlist(allotherdaysdiff)


```

```{r ttest, include=F, message = F}
# test if distance to next day is less than distance to all other days
wilcox.test(x, y, paired = T)

```


```{r microbiome_distance_test, echo = F, fig.height=2.5, fig.width=2.5}

#plot

x <- as.data.frame(x)
colnames(x) <- "distance"
x$mb_dist_to <- "To next day"

y <- as.data.frame(y)
colnames(y) <- "distance"
y$mb_dist_to <- "To other days"

plot <- rbind(x,y)
myplot <- ggplot(data = plot, aes(x = mb_dist_to, y = distance)) + 
  geom_jitter(alpha = 0.7, width = 0.35, color = "#5f86b7", size = 2) +
  geom_boxplot(outlier.shape = NA, fill = NA) +
  annotate("text", x = 1.5, y = 25, label = "italic(p) < 0.001", size = 3, parse = T) +
  theme_classic() +
  xlab("Distance from today") +
  ylab("Mean Aitchison's distance")+
  theme(axis.text = element_text(size = 7),
        axis.title = element_text(size =7))
  
  
myplot

```



# We have seen that diet on the weekend is more variable than on the weekdays. 
# So, we asked, is the microbiome more variable on Monday-Tuesday (presumably in response to changes in diet on Fri-Sat-Sun)



```{r weekday v weekend, include =F}

set.seed(42)

weekday <- NULL
weekend <- NULL
weekdaymean <- NULL
weekendmean <- NULL
for (i in 1:length(list.mbdist)) {
  # subset to just one person 
  x <- as.matrix(list.mbdist[[i]])
  
  # rename distance matrix with days of the week
  weekdays <- map[map$X.SampleID %in% rownames(x),]$Sample.Week.Day
  rownames(x) <- weekdays
  colnames(x) <- weekdays
  # melt 
  melt_dist <- melt(x)
  
   # extract the values for Mon v. Tues, Tues v. Wed, Wed v. Thurs, Thurs v. Fri as weekday
  weekend[[i]] <- melt_dist[melt_dist$Var1 == "Monday" & melt_dist$Var2 == "Tuesday",]$value
    
  
  # extract the values for Fri v. Sat, Sat v. Sun as weekend day
  weekday[[i]] <- melt_dist[melt_dist$Var1 == "Wednesday" & melt_dist$Var2 == "Thursday"|
                              melt_dist$Var1 == "Thursday" & melt_dist$Var2 == "Friday" |
                              melt_dist$Var1 == "Saturday" & melt_dist$Var2 == "Sunday" |
                              melt_dist$Var1 == "Sunday" & melt_dist$Var1 == "Monday",]$value
  
  #weekday[[i]] <- sample(weekday[[i]], length(weekend[[i]]))
  
  weekdaymean[[i]] <- mean(weekday[[i]])
  weekendmean[[i]] <- mean(weekend[[i]])

  
}

x <- unlist(weekendmean)
y <- unlist(weekdaymean)

names(x) <- unique(map$UserName)
names(y) <- unique(map$UserName)

x <- x[!names(x) %in% c("MCTs11", "MCTs12")]
y <- y[!names(y) %in% c("MCTs11", "MCTs12")]



```


```{r ttest2, include=F, message = F}
# is the Monday - Wed microbimoe (weekend) more variable than Wed-Friday (weekday) microbiome?

wilcox.test(x, y, paired = T)

```


```{r mb_weekday_v_weekend, echo = F, fig.height=2.5, fig.width=2.5}

#plot
pval <- wilcox.test(x, y, paired = T)$p.value

plotx <- as.data.frame(x)
colnames(plotx) <- "distance"
plotx$distance_between <- "Between\nweekend\naffected days"

ploty <- as.data.frame(y)
colnames(ploty) <- "distance"
ploty$distance_between <- "Between\nweekday\naffected days"

plot <- rbind(plotx,ploty)
myplot <- ggplot(data = plot, aes(x = distance_between, y = 1/distance)) + 
  geom_jitter(alpha = 0.7, width = 0.35, color = "#5f86b7", size = 2) +
  geom_boxplot(outlier.shape = NA, fill = NA) +
  #annotate("text", x = 1.5, y = 27, label = paste0("italic(p) ==",round(pval,2)), size = 3, parse = T) +
  theme_classic() +
  ylab("Aitchison's distance") +
  theme(axis.text = element_text(size = 7),
        axis.title = element_text(size =9),
        axis.title.x = element_blank())
  


myplot

```

```{r soylent v others, include =F}
# TODO: ALTER FOR SOYLENT!


set.seed(42)

names(list.mbdist) <- unique(map$UserName)

soylent <- list.mbdist[names(list.mbdist) %in% c("MCTs11", "MCTs12")]
others <- list.mbdist[!names(list.mbdist) %in% c("MCTs11", "MCTs12")]

soylent <- lapply(soylent, function(x) rowMeans(x))
soylent <- lapply(soylent, function(x) mean(1/x))

others <- lapply(others, function(x) rowMeans(x))
others <- lapply(others, function(x) mean(1/x))

x <- unlist(soylent)
y <- unlist(others)



```


```{r ttest3, include=F, message = F}
# is the Monday - Wed microbimoe (weekend) more variable than Wed-Friday (weekday) microbiome?

wilcox.test(x, y)

```

```{r mb_soylent_v_others, echo = F, fig.height=2.5, fig.width=2.5}

#plot
pval <- wilcox.test(x, y)$p.value

plotx <- as.data.frame(x)
colnames(plotx) <- "distance"
plotx$distance_between <- "Meal replacement"

ploty <- as.data.frame(y)
colnames(ploty) <- "distance"
ploty$distance_between <- "Habitual diet"

plot <- rbind(plotx,ploty)
myplot <- ggplot(data = plot, aes(x = distance_between, y = distance)) + 
  geom_jitter(alpha = 0.7, width = 0.35, color = "#5f86b7", size = 2) +
  geom_boxplot(outlier.shape = NA, fill = NA) +
  #annotate("text", x = 1.5, y = 0.05, label = paste0("italic(p) ==",round(pval,3)), size = 3, parse = T) +
  theme_classic() +
  ylab("Microbiome Stability") +
  theme(axis.text = element_text(size = 7),
        axis.title = element_text(size =7),
        axis.title.x = element_blank())
  


myplot

```
```{r plot distancetonextday, fig.height = 2, fig.width= 3}

names(day1diffall) <- unique(map$UserName)
mydistance <- stack(day1diffall)
names(weekdaysall) <- unique(map$UserName)
mydays <- stack(weekdaysall)


plot <- cbind(mydistance, mydays)
colnames(plot) <- c("Distance", "UserName", "Day", "UserName2")

plot$Diet <- ifelse(plot$UserName %in% c("MCTs11", "MCTs12"), "Shake Drinkers", "Normal Diet")

plot <- subset(plot, !plot$UserName %in% c("MCTs06", "MCTs29"))
plot$Day <- as.factor(plot$Day)


plot_mean<- aggregate(plot$Distance, by = list(plot$Day, plot$Diet), FUN = mean)
colnames(plot_mean) <- c("Day", "Mean Diet", "Distance")


ggplot(plot, aes(x = Day, y = Distance)) + 
  geom_line(aes(group = UserName), color = "grey", aes = 0.75) +
  geom_line(data = subset(plot, plot$UserName == "MCTs11"), aes(group = UserName), color = "#5f86b7", size = 1) +
  geom_line(data = subset(plot, plot$UserName == "MCTs12"), aes(group = UserName), color = "#5f86b7", size = 1) +
  #geom_smooth(data = plot_mean, aes(group = `Mean Diet`, color = `Mean Diet`)) +
  ylab("Microbiome distance\nfrom previous day") +
  xlab("Study Day") +
  theme_classic() +
  theme(axis.title = element_text(size = 8))
  



```


